---
title: "Getting Started with courieR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with courieR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

courieR moves your installed R packages from an old R version to a new one on the same machine. This vignette walks you through both the dashboard and the programmatic CLI.

## Installation

Install courieR on your **new** R version:

```{r}
remotes::install_github("lennon-li/courieR")
```

You do not need to install anything on your old R version.

## Step 1 — Discover Your R Installations

`find_routes()` scans the system and returns a data frame of every R version it finds:

```{r}
library(courieR)

routes <- find_routes()
routes[, c("version", "rscript_path", "is_current")]
#>   version                              rscript_path is_current
#> 1   4.2.3  C:/Program Files/R/R-4.2.3/bin/Rscript.exe      FALSE
#> 2   4.4.1  C:/Program Files/R/R-4.4.1/bin/Rscript.exe       TRUE
```

`is_current = TRUE` marks the R you are running right now — that is your destination.

## Step 2a — Use the Dashboard (Recommended)

Launch the Shiny hub:

```{r}
open_hub()
```

The dashboard opens in your browser with two tabs:

**Migrate tab** — the main workflow:

1. Select your source (old R) from the From dropdown
2. Select your destination (new R) from the To dropdown — defaults to current R
3. Review the package checklist — all packages are selected by default
4. Click **Deliver** and watch the progress output

**Advanced tab** — for inspection and diagnostics:

- Manifest: full package list for any R version
- Delivery Receipt: results of the last ship
- Diagnostics: raw logs if anything went wrong

## Step 2b — Use the CLI

Prefer code? `ship()` does the same thing programmatically:

```{r}
# always do a dry run first to see the plan
result <- ship(
  source_path = routes$rscript_path[1],   # old R
  target_path = routes$rscript_path[2],   # new R (current)
  dry_run = TRUE
)

# review what will be installed or upgraded
print(result$plan[, c("package", "action", "version.x")])
```

When you are happy with the plan, run for real:

```{r}
result <- ship(
  source_path = routes$rscript_path[1],
  target_path = routes$rscript_path[2]
)

# check results
result$results[result$results$status == "error", ]
```

## Inspecting Libraries Manually

`manifest()` lists packages for any R version. `inventory()` compares two manifests:

```{r}
old_pkgs <- manifest(rscript_path = routes$rscript_path[1])
new_pkgs <- manifest(rscript_path = routes$rscript_path[2])

comp <- inventory(old_pkgs, new_pkgs)
nrow(comp$missing)    # packages in old R but not new R
nrow(comp$outdated)   # packages where old R has a newer version
```

## Tips

- courieR skips base and recommended packages — only user-installed packages are migrated
- If a package fails to install, check `result$results` for the error message
- CRAN packages are reinstalled fresh from CRAN; GitHub packages are detected and sourced via pak
- Run `find_routes(search_paths = "C:/custom/R")` to include non-standard install locations
